---
title: "Comprehensive Data Analysis and Predictive Modeling Capstone Report"
subtitle: "Optimizing Retail Operations, Footfall Dynamics, and Revenue Drivers Across NYC Coffee Branches"
author: "Data Analyst Intern"
date: "`r Sys.Date()`"
output: 
  word_document:
    toc: true
    toc_depth: 3
    fig_width: 6.5
    fig_height: 4.0
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(readxl)
library(tidyverse)
library(lubridate)
library(fastDummies)
library(caret)
library(scales)
library(gridExtra)
library(knitr)
```

# 1. Executive Summary & Project Introduction

This capstone project evaluates transactional sales records from a multi-location specialty retail coffee enterprise operating across New York City (Astoria, Hell's Kitchen, and Lower Manhattan) from January 1, 2023, through June 30, 2023. The scope includes 149,116 transaction records generating $698,812.33 in gross sales.

```{r load_data}
df_raw <- read_excel("Coffee Shop Sales.xlsx", sheet = "Transactions")
dim(df_raw)
colSums(is.na(df_raw))
```

# 2. Data Preparation, Cleaning & Preprocessing

```{r data_preprocessing}
df_clean <- df_raw %>%
  mutate(
    transaction_date = as.Date(transaction_date),
    transaction_hour = hour(transaction_time),
    day_of_week      = factor(weekdays(transaction_date), 
                              levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")),
    month_name       = factor(months(transaction_date), 
                              levels = c("January", "February", "March", "April", "May", "June")),
    total_sales      = transaction_qty * unit_price,
    store_location   = as.factor(store_location),
    product_category = as.factor(product_category),
    product_type     = as.factor(product_type),
    is_weekend       = ifelse(day_of_week %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
  )

# Tukey's IQR Winsorization
cap_iqr <- function(x) {
  q <- quantile(x, probs = c(0.25, 0.75), na.rm = TRUE)
  iqr_val <- q[2] - q[1]
  lower_b <- max(0, q[1] - 1.5 * iqr_val)
  upper_b <- q[2] + 1.5 * iqr_val
  pmax(pmin(x, upper_b), lower_b)
}

df_clean$unit_price_capped  <- cap_iqr(df_clean$unit_price)
df_clean$total_sales_capped <- cap_iqr(df_clean$total_sales)

# Before vs After Summary Table
kable(rbind(
  "Original Unit Price" = summary(df_clean$unit_price),
  "Capped Unit Price"   = summary(df_clean$unit_price_capped),
  "Original Total Sales"= summary(df_clean$total_sales),
  "Capped Total Sales"  = summary(df_clean$total_sales_capped)
), caption = "Outlier Treatment via Tukey's IQR Winsorization")
```

```{r plot_outliers, fig.cap="Figure 1: Before and After Outlier Treatment Boxplots"}
p1 <- ggplot(df_clean, aes(y = unit_price)) +
  geom_boxplot(fill = "#93C5FD", outlier.color = "red", alpha = 0.8) +
  theme_minimal() + labs(title = "A: Raw Unit Price", y = "Price ($)")

p2 <- ggplot(df_clean, aes(y = unit_price_capped)) +
  geom_boxplot(fill = "#6EE7B7", alpha = 0.8) +
  theme_minimal() + labs(title = "B: Winsorized Unit Price", y = "Price ($)")

grid.arrange(p1, p2, ncol = 2)
```

# 3. Exploratory Data Visualization & Business Insights

```{r plot_hourly, fig.cap="Figure 2: Hourly Transaction Footfall by Branch"}
ggplot(df_clean, aes(x = factor(transaction_hour), fill = store_location)) +
  geom_bar(position = position_dodge(0.8), width = 0.75, alpha = 0.9) +
  scale_fill_manual(values = c("Astoria" = "#3B82F6", "Hell's Kitchen" = "#10B981", "Lower Manhattan" = "#F59E0B")) +
  scale_y_continuous(labels = comma_format()) +
  theme_minimal() +
  labs(title = "Hourly Footfall Distribution by Store Location", x = "Hour (24-Hr Clock)", y = "Orders", fill = "Store")
```

```{r plot_monthly, fig.cap="Figure 3: Monthly Revenue Growth (Jan - Jun 2023)"}
monthly_summary <- df_clean %>%
  group_by(month_name) %>%
  summarise(total_rev = sum(total_sales), .groups = "drop")

ggplot(monthly_summary, aes(x = month_name, y = total_rev / 1000, group = 1)) +
  geom_line(color = "#1E40AF", linewidth = 1.3) +
  geom_point(color = "#1E40AF", fill = "#60A5FA", shape = 21, size = 4) +
  scale_y_continuous(labels = function(x) paste0("$", x, "k"), limits = c(60, 180)) +
  theme_minimal() +
  labs(title = "Monthly Revenue Growth Trajectory", x = "Month", y = "Gross Revenue ($k)")
```

```{r plot_category, fig.cap="Figure 4: Gross Revenue by Product Category"}
category_summary <- df_clean %>%
  group_by(product_category) %>%
  summarise(total_rev = sum(total_sales), .groups = "drop")

ggplot(category_summary, aes(x = reorder(product_category, total_rev), y = total_rev / 1000)) +
  geom_col(fill = "#0D9488", width = 0.65) +
  coord_flip() +
  scale_y_continuous(labels = function(x) paste0("$", x, "k")) +
  theme_minimal() +
  labs(title = "Gross Revenue by Category", x = "Category", y = "Revenue ($k)")
```

# 4. Inferential Statistical Analysis & Predictive Modeling

```{r hypothesis_testing}
# 1. One-Way ANOVA across Store Locations
anova_fit <- aov(total_sales_capped ~ store_location, data = df_clean)
summary(anova_fit)

# 2. Two-Sample t-Test
t.test(total_sales_capped ~ is_weekend, data = df_clean, var.equal = FALSE)
```

```{r predictive_modeling}
# 80/20 Train-Test Split
set.seed(42)
train_idx <- createDataPartition(df_clean$total_sales_capped, p = 0.80, list = FALSE)
train_df  <- df_clean[train_idx, ]
test_df   <- df_clean[-train_idx, ]

# Multiple Linear Regression
model_lm <- lm(
  total_sales_capped ~ unit_price_capped + transaction_hour + store_location + product_category,
  data = train_df
)
summary(model_lm)

# Out-of-Sample Performance
preds <- predict(model_lm, newdata = test_df)
eval_df <- data.frame(
  RMSE = RMSE(preds, test_df$total_sales_capped),
  MAE  = MAE(preds, test_df$total_sales_capped),
  R2   = R2(preds, test_df$total_sales_capped)
)
kable(eval_df, caption = "Holdout Test Set Performance Metrics")
```

# 5. Strategic Recommendations & Conclusion

1. **Morning Labor Allocation:** Realign shift schedules to assign 60% of frontline baristas between 7:00 AM and 11:00 AM to handle the 42.6% morning volume concentration.
2. **Breakfast Bundling Programs:** Launch bundled beverage + pastry promotions to increase multi-item order rates above the current 1.44 unit average.
3. **Targeted Merchandising in Lower Manhattan:** Capitalize on Lower Manhattan's higher average ticket size ($4.81 vs $4.59 baseline) with packaged whole bean and catering displays.

